---
title: "Assgt Wk 2 RR Course"
author: "Bryan K"
date: "December 7, 2017"
output: 
  md_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Activity Monitoring Devices

Read the raw data from a local csv file and add to r. Then set up some values for the code and set up the dates column.

```{r activity, warning=FALSE, message=FALSE}
library(dplyr)
library(magrittr)
library(data.table)
library(knitr)
library(kableExtra)
library(lattice)

activity <- read.csv("C:/Users/Bryan/Desktop/DATA SCIENCE CERT/Coursera5_ReproducibleResearch/Ass2/activity.csv", header=TRUE, stringsAsFactors = FALSE)
activity$date <- as.Date(activity$date)

nactivity <- activity  ## Retain NA values for later calculations
activity <- na.omit(activity)  ## Remove NA values for parts 1,2
```

Take the number of steps from the device, gather the total, and return a histogram of the total steps per day.  This will then be summarized as a table showing the mean and median for the steps/day:

```{r total steps, tidy = TRUE }
options(digits = 4)
opts_chunk$set(tidy.opts=list(width.cutoff=80),tidy=TRUE)
tot.step <- aggregate(steps ~ date,FUN=sum,data=activity)
tot.step$date <- as.Date(tot.step$date, format="%Y-%m-%d")
names(tot.step) <- c("Date","TotalSteps")

hist(tot.step$TotalSteps,    #Step histogram as Total/day
          main="Total Steps per Day",
          xlab = "Step Range", col = 3,
          breaks=20) 
          
# Follow the histogram with statistical summary functions:
mean(tot.step$TotalSteps)

median(tot.step$TotalSteps)
```
Create a table to examine the data array into a data frame processing the mean steps taken for each 5-min interval in a 24 hour period. Begin by calculating the mean steps of each day via 5-min intervals, then turn it into a (readable) table and xy plot:
```{r steps per interval}
time.step <- as.data.frame(tapply(activity$steps,activity$interval,mean,simplify=FALSE)) # mean calculation
time.step <- cbind(rownames(time.step),time.step)
colnames(time.step) <- c("interval", "mean.step")
rownames(time.step) <- c()
time.step$mean.step <- as.numeric(as.character(time.step$mean.step))
time.step$interval <- as.numeric(as.character(time.step$interval))
xyplot(mean.step~interval, #Generate a time-based xy plot with the interval data
       data=time.step,type = "l",
       main = "Average Steps per Day",
       xlab = "Time Interval (min)", ylab = "Mean Steps",
       scales=list(tck=c(1,0))
       ) 
```

```{r, echo = FALSE} 
max.steps <- max(as.numeric(as.character(time.step$mean.step)))
max.int <- time.step$interval[which(as.numeric(as.character(time.step$mean.step)) == max.steps)]
```
From this data, we can see that the max avg. steps is `r max.steps`, which occurs at the interval `r max.int`.

We should now take a look at the values that have been left OUT of the data and determine what to do with them.  Let's make a new column counting the NA values in the total data, 'nactivity'.  
```{r}
nactivity$na_count <- apply(nactivity, 1, function(x) sum(is.na(x)))
```
```{r, echo = FALSE}
tot.na <- sum(nactivity$na_count)
tot.data <- nrow(nactivity)
```
We now have a new column that effectivly gives a T/F value for NAs which reports a sum of `r tot.na` rows of NA data out of `r tot.data`.

To fill in the NA values, we will use the mean for that five-minute interval:
```{r estimate NA values}
tot.activity <- nactivity
colnames(tot.activity)
for (i in 1:nrow(tot.activity)) {
      if (is.na(tot.activity$steps[i])) {
            this.step <- as.numeric(time.step$mean.step[which(time.step$interval == tot.activity$interval[i])])
            tot.activity$steps[i] <- this.step
      }
}

tot.step <- aggregate(steps ~ date,FUN=sum,data=tot.activity)
tot.step$date <- as.Date(tot.step$date, format="%Y-%m-%d")
names(tot.step) <- c("Date","TotalSteps")

hist(tot.step$TotalSteps,    #Step histogram as Total/day
          main="Completed Total Steps per Day",
          xlab = "Step Range", col = 2,
          breaks=20) 
          
# Follow the histogram with statistical summary functions:
mean(tot.step$TotalSteps)

median(tot.step$TotalSteps)

```

It is likely that the steps measured will vary from weekday to weekend.  Herein, we shall add columns to the dataset with the weekday and a T/F column determining if it is a weekend.

```{r interval measurement steps}
tot.activity$week <- weekdays(tot.activity$date)
tot.activity$weekend <- tot.activity$week == "Saturday" | tot.activity$week == "Sunday"

#Interval steps for a weekday
time.step.wd <- tapply(tot.activity$steps[which(tot.activity$weekend== FALSE)],tot.activity$interval[which(tot.activity$weekend== FALSE)],mean,simplify=FALSE)
time.step.wd <- as.data.frame(time.step.wd)
time.step.wd$weekend <- FALSE
setDT(time.step.wd,keep.rownames = TRUE)[]
colnames(time.step.wd) <- c("Interval","Mean","Weekend")

#Interval steps for a weekend
time.step.we <- tapply(tot.activity$steps[which(tot.activity$weekend== TRUE)],tot.activity$interval[which(tot.activity$weekend== TRUE)],mean,simplify=FALSE)
      time.step.we <- as.data.frame(time.step.we)
      time.step.we$weekend <-TRUE
      
      setDT(time.step.we,keep.rownames = TRUE)[]
      colnames(time.step.we) <- c("Interval","Mean","Weekend")

#Generate a numeric table & plot for weekday/weekend, now merged
time.step.week <- rbind(time.step.wd, time.step.we)
      colnames(time.step.week) <- c("interval","mean","weekend")
      time.step.week$mean <- as.numeric(time.step.week$mean)
      time.step.week$interval <- as.numeric(time.step.week$interval)

xyplot(mean~interval|weekend # interval vs. mean as a factor of weekend (T/F)
       ,data=time.step.week,
       type = "l", main = "NA-Free Average Steps per Day",
       xlab = "Time Interval (min)", ylab = "Mean Steps",
       strip = strip.custom(factor.levels = c("Weekday","Weekend"), 
                  par.strip.text = list(cex = 0.8) ),
       layout=c(1,2))
```

One can see from the above graphs that this client has a much higher, but consistent, step count on weekends averaging around 125 steps.  During a weekday, there is a high quantity of steps (230) around 8am when the person is likely scrambling to get to work.  This is followed by a daily low of steps, presumably due to a stationary 10 - 6 job.
---
